Dissertation Proposal

### 2023-10-31 proposal.tex
plan for RQ1

\usepackage[utf8]{inputenc} % Use UTF-8 encoding
\usepackage[T1]{fontenc} % Use the T1 font encoding
\usepackage{geometry} % Adjust page margins
\usepackage{amsmath, amssymb} % For math symbols and equations
\usepackage{graphicx} % For including images
\usepackage{titlesec} % For customizing section titles
\usepackage[style=authoryear]{biblatex}  % For managing references
\addbibresource{bibfile.bib}
\usepackage{hyperref}
\usepackage[colorinlistoftodos]{todonotes}
\newcommand{\miscon}[1]{\textbf{#1}}
\newcommand{\surmisedMiscon}[1]{\miscon{#1}$^\ddag$}
\newcommand{\pValue}[0]{\textit{p}\nobreakdash-value}

@@

\begin{description}
  \item[RQ1] What are the misconceptions that are about program
  behavior, apply to most modern programming languages, and present
  (even) in students with prior programming backgrounds?
  \item[RQ2] What would be a tutoring system (the SMoL Tutor) that
  draws on cognitive and educational psychology concepts, utilizes a
  notional machine (the Stacker), and aims at correcting the
  misconceptions?
  \item[RQ3] How effective is SMoL Tutor at correcting the
  misconceptions?
\end{description}

@@

@@

@@

By the end of Fall 2023, I would like to improve the SMoL Tutor:
\begin{enumerate}
  \item Ask students to explained when they gave a wrong answer in a
  interepreting question. With student explanations, I check whether
  choosing the wrong answer indicates holding the corresponding misconception(s).
  \item Adjust the collection of interpreting questions such that
  every misconception has at least two questions that aims at
  correcting the misconception.
  \item Partially randomize the order of questions.
  \item Try to reduce the chance of missing values.
\end{enumerate}

>>>>

\usepackage[utf8]{inputenc} % Use UTF-8 encoding
\usepackage[T1]{fontenc} % Use the T1 font encoding
\usepackage{geometry} % Adjust page margins
\usepackage{amsmath, amssymb} % For math symbols and equations
\usepackage{graphicx} % For including images
\usepackage{titlesec} % For customizing section titles
\usepackage[style=authoryear]{biblatex}  % For managing references
\addbibresource{bibfile.bib}
\usepackage{hyperref}
\usepackage[colorinlistoftodos]{todonotes}
\newcommand{\miscon}[1]{\textbf{#1}}
\newcommand{\surmisedMiscon}[1]{\miscon{#1}$^\ddag$}
\newcommand{\pValue}[0]{\textit{p}\nobreakdash-value}
\newcommand{\rqI}{What are the misconceptions that are about program behavior, apply to most modern programming languages, and present (even) in students with prior programming backgrounds?}
\newcommand{\rqII}{What would be a tutoring system that aims at correcting the misconceptions, draws on cognitive and educational psychology concepts, utilizes a notional machine, and seems to be effective?}

@@

\begin{description}
  \item[RQ1] \rqI
  \item[RQ2] \rqII
\end{description}

@@

  \item I present a notional machine, the Stacker, that is used by the
  SMoL Tutor to explain programs and might be useful in itsown.

@@

\section{Background: Misconceptions}%
\label{s:background-misconceptions}

\section{Background: Program Behavior}

\section{Plan for \textbf{RQ1}}%
\label{s:methodology-rq1}

\begin{description}
  \item[RQ1] \rqI
\end{description}

To ensure that I am studying program behavior rather than some
implementation details, I plan to focus on the results of running
programs. I approach the second predicate by modeling the
``consensus'' of modern programming language with SMoL. Because SMoL
represents the semantic core, misconceptions about SMoL presumably
also apply to most modern PLs. I approach the third and last predicate
by collecting data in population that very likely have prior
experience. For example, my priliminary results are build upon data
collected in a upper-year programming language course.

In order to find a reasonably comprehensive collection of
misconceptions, I seek for inputs from students. Most prior works
(\Cref{s:background-misconceptions}) have experts coming up with a
list of potential misconceptions and then filter the potential
misconceptions with student inputs. This approach can't possibly find
misconceptions not considered by experts and hence is necessarily
affected by experts' blind spots. I plan to overcome experts'
blind spots by avoid presuming misconceptions. Concretely, my first
step was a two-year process to find (a) student-written programs that
tend to trip up students and (b) incorrect responses to those programs
(\Cref{s:gen-prob-quizius}), using a tool for the purpose
(\Cref{s:quizius}).

Another source of misconceptions are student responses in the tutoring
system. The tutoring system, which I will explain in details when I
talk about RQ2 (\Cref{s:methodology-rq2}), includes many
multiple-choice questions (MCQs) that asks students to predict the
result(s) of running a program. MCQ is a commonly used task format for
detecting misconceptions [CITE]. MCQs in their own only detect
\emph{patterns of errors}. To detect \emph{misconceptions} -- student
conceptions that produces pattern of errors, I need student-written
explanations for their wrong answers. To this end, I plan to ask a
follow up questions after the MCQs: when students make a wrong choice,
the follow-up asks \emph{why} they made the choice.

I expect to find multiple misconceptions. Each misconception is
represented by a collection of program-result-explanation triples.
There should be a small number of programs. For each of these
programs, a wrong result is commonly predicted by students from
various population. Furthermore, these students give similar explanations
for all the wrong answers. I also would like to provide a definitional
interpreter for the misconception.

For example, I might find a misconception called \textbf{DefByRef}:
when students are asked to predict the result of
\begin{lstlisting}
  (defvar x 12)
  (defvar y x)
  (set! y 0)
  x
  y
\end{lstlisting}
many of them responded \lstinline|12 0|; and when asked to predict
\begin{lstlisting}
  (defvar x 12)
  (defvar y x)
  (set! x 0)
  x
  y
\end{lstlisting}
many of them responded \lstinline|0 12|; and these students all
explained something like ``\lstinline|y| is \lstinline|x|''. To
represent this misconception, I provide a interpreter that almost
works like the referential interperter of SMoL, but \lstinline|defvar|
aliases variables.

\section{Plan for \textbf{RQ2}}%
\label{s:methodology-rq2}

\begin{description}
  \item[RQ2] \rqII
\end{description}

Prior works suggest that presenting refutation text and encouraging
students to compare cases are helpful for learning, and that notional
machines are useful for explaning program behavior.

I have built a tutor that incorporate these ideas
(\Cref{s:system-tutor}).

This tutor uses a lot of MCQs illustrated in
\Cref{f:smol-tutor-example}.

@@

By the end of Fall 2023, I would like to improve the SMoL Tutor:
\begin{enumerate}
  \item Ask students to explained when they gave a wrong answer in a
  interepreting question. With student explanations, I check whether
  choosing the wrong answer indicates holding the corresponding misconception(s).
  \item Adjust the collection of interpreting questions such that
  every misconception has at least two questions that aims at
  correcting the misconception.
  \item Partially randomize the order of questions.
\end{enumerate}
